---
title: "vcr_fish_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vcr_fish_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(lterdatasets)
library(vcrfish)
library(tidyverse)
```

An initial cleaning of the VCR fish dataset:
```{r}
vcr_fish <- vcr_fish %>%
  mutate( sample_date=as.Date(sample_date, "%Y-%m-%d") ) %>%
  mutate(sample_year=as.integer( format(sample_date, format = "%Y") ),
         sample_month=as.integer( format(sample_date, format = "%m") ),
         sample_day=as.integer( format(sample_date, format = "%d") )
         )
vcr_fish <- vcr_fish %>%
  mutate(sample_hour=as.integer(
              format(as.POSIXct(sample_time, format = "%H:%M"), format = "%H")
              ),
         sample_min=as.integer(
           format(as.POSIXct(sample_time, format = "%H:%M"), format = "%M")
           )
        )
```

Let's take a look at the variables within this dataset:
```{r}
vcr_fish %>% head()
sapply(vcr_fish, typeof)
```

Let's take a closer look at differences between fish species to learn more.
```{r}
unique(vcr_fish$species_name)
```

In the code below, I am filtering the fish species type to have greater than 200 samples to ensure that there is a sufficient sample size in our analyses later on.
```{r}
# Count the number of names 
names_count <- table(vcr_fish$species_name)

# Subset the orginial dataset with only the species names specified by the line above.
vcr_fish2 <- subset(vcr_fish, species_name %in% names(names_count[names_count > 200]))
```

Now, we have 5 different fish species (pipefish, silversides, pinfish, silver perch, and bay anchovy) with more than 200 rows of data.
```{r}
unique(vcr_fish2$species_name)
```

A five number summary is a statistical summary that includes the minimum, maximum, median, mean, first quartile, and third quartile values of a given dataset. Let's calculate the five number summary for the fish length of all the species sampled.
```{r}
vcr_five_number_sum <- summary(vcr_fish2$length)
vcr_five_number_sum
```
Upon a closer look, we see that our IQR = 95 - 50 (Q3 - Q1) = 45 inches, where 50% of the data sits inside these two quartiles and 50% sits outside of these quartiles. This is a large disparity, so lets create five number summaries for each fish species individually.

Let's view the five number summary for all 5 fish species.
```{r}
# Create separate datasets for all 5 species
pipefish <- vcr_fish2 %>% filter(species_name=="Pipefish")
silversides <- vcr_fish2 %>% filter(species_name=="Silversides")
pinfish <- vcr_fish2 %>% filter(species_name=="Pinfish")
silverperch <- vcr_fish2 %>% filter(species_name=="Silver Perch")
bayanchovy <- vcr_fish2 %>% filter(species_name=="Bay Anchovy")

# Individual five number summaries
pipefish_sum <- summary(pipefish$length)
silversides_sum <- summary(silversides$length)
pinfish_sum <- summary(pinfish$length)
silverperch_sum <- summary(silverperch$length)
bayanchovy_sum <- summary(bayanchovy$length)

pipefish_sum
silversides_sum
pinfish_sum
silverperch_sum
bayanchovy_sum
```

Boxplots are a graphical and visual way to display numerical data through four quartiles. This visualization is advantageous because it easily shows a given variable's variability, skewness, outliers, and symmetry. Additionally, boxplots help to visualize the five number summary that we calculated above.

Let's graph a boxplot of the filtered fish data to look at initial trends in the data.
```{r}
ggplot(data = vcr_fish2, aes(x=species_name, y=length)) +
  geom_boxplot()
```
It appears that the pinfish species has an outstanding outlier. Let's go ahead and remove this outlier just so that we can view the boxplots at a closer level.
```{r}
pinfish_outlier <- vcr_fish2 %>% slice_max(length) #store it in case
vcr_fish3 <- vcr_fish2 %>% filter(length!=pinfish_outlier$length)
```

Let's add color to our boxplots to differentiate fish species from each other.
```{r}
ggplot(data = vcr_fish3, aes(x=species_name, y=length, color=species_name)) + 
  geom_boxplot()
```

Additionally, we can add a main title and alter the axis' titles.
```{r}
ggplot(data = vcr_fish3, aes(x=species_name, y=length, color=species_name)) + 
  geom_boxplot() +
  labs(title = "Boxplots of VCR Fish Species",
       x = "Species Name",
       y = "Length") + 
  theme(plot.title = element_text(hjust = 0.5)) 
```
Overall, we see that pipefish as a whole species are longer than the other fish species given they have the highest median value around 140.

